Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-05-01 paper/tidy-missing-data-paper.Rmd
tighten up a few sections and tighten up the use of mentions to shadow matrix, flag  few more things to use in the section

To explore missing data there needs to be a robust representation of missing
values that integrates into analysis pipelines. This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis. The $B$ matrix, where 0 and 1 indicate not missing and missing (respectively), provides a useful template, as it is very common in missing data literature.  This matrix was used to explore missing values in
the interactive graphics library XGobi, called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness. The shadow matrix could be interactively linked to the data. There are some limitations to the shadow matrix, however. Namely, the values 0 and 1 can be confusing representations of missing values - does 0 indicate an absence of observation, or the presence of a missing value?

@@

1. **Missing value labels**. Simple labels, "NA" and "!NA", for missing and not missing, clearly identifies missing values for analysis and plotting. This is an improvement over the commonly used binary matrix ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value. Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3). (principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this. Anybody looking at these values)

1. **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)). The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)). Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)). {Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.  SAS, SPSS, and STATA all implement } {The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state. Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.} {(discuss the principles, rather than what we've done)}. {(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)}.

@@

(change cohesiveness to connectedness)

@@

The acronym MANET [@Unwin1996] stands for "Missings Are Now Equally Treated". Missing values are often ignored when plotting data, but  plots can help to reveal the type of missing values patterns existing in the data, and even help to assess whether the fall into the MCAR, MAR or even MNAR labels.

Here we summarise systematically exploring the missing patterns visually, and define  useful plots to make relative to the nabular  data structure.

<!--
(note: next section to work on)

note: First sentence needs to be more on track with the overall summary of what this section is about...

- Graphics are important to understand the missingness structure.
- The multivariate relationships
- Working through a hierarchy of how to explore missings - from overview to 
- Sentence linking the two.
- How to visualise a binary matrix of missing vs not, and the dependencies between those binary variables as well.
- It is more than that, when we visualise the complete cases, we don't want to lose track of the missings themselves. This comes back to the data structure
- gg_miss_var and gg_miss_upset and vis_miss are all showing how we can summarise the missings.
- but the addition of missing counts into a histogram, allows us to understand that there are many missings - and we have a skewed distribution o Ozone. But we want to know where these values might have been located. tying this back to the nabular structure, and also the terminology of MCAR MAR, etc. (We would expect if data were MCAR for these points to be representatively thoughout the histogram)

When we define the graphics, describe which part of the nabular matrix we are going to display (overview - just binary parts), the summaries, are summaries of the binary, the histogram and points are both the missings and the complete values.
-

- Combining that with the data itself, looking at the patterns in the complete cases, 
- from the overview through to the univariate

Visualizing missingness in a dataset is essential to understand its structure. This section discusses four different areas of visualization of missing data: overviews (\@ref(overviews)), univariate (\@ref(univariate)), bivariate (\@ref(bivariate)), and multivariate (\@ref(multivariate)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance.
-->

@@

@@

Both variables and cases can be displayed using a heatmap style visualization,available with `vis_miss()` from the \pkg{visdat} package [@visdat]. This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:vis-miss)). The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B). Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation. Using \pkg{ggplot2} makes the plot easily customizable. This also uses the shadow matrix as the basis for the plot.

@@

@@

>>>>

If a data structure is to be added to facilitate exploring missing data, it needs to be simple to understand and transport with a data analysis. A useful template is the $B$ matrix, where 0 and 1 indicate not missing and missing (respectively), as it is very common in missing data literature.  This matrix was used to explore missing values in
the interactive graphics library XGobi, called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness. The shadow matrix could be interactively linked to the data.

There are some limitations to the shadow matrix, however. Namely, the values 0 and 1 can be confusing representations of missing values - does 0 indicate an absence of observation, or the presence of a missing value?

@@

1. **Missing value labels**. Simple labels, "NA" and "!NA", for missing and not missing, clearly identifies missing values for analysis and plotting. This is an improvement over ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value. Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).

(NOTE: principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this. Anybody looking at these values)

1. **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)). The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)). Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)). {Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.  SAS, SPSS, and STATA all implement }

{NOTE: The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state. Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.}

{NOTE: (discuss the principles, rather than what we've done)}. {(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)}.

@@

(NOTE: change cohesiveness to connectedness)

@@

Missing values are often ignored when plotting data, but plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or even MNAR.
Here we summarise systematically exploring the missing patterns visually, and define useful plots to make relative to the nabular data structure, working from coarse to fine levels of granularity in understanding missingness.

- The acronym MANET [@Unwin1996] stands for "Missings Are Now Equally Treated".

@@

- How to visualise a binary matrix of missing vs not, and the dependencies between those binary variables as well.
- These overview plots are looking at the shadow matrix, and the summaries of each of the variables
- In this section we  describe which part of the nabular matrix we are going to display (overview - just binary parts), the summaries are summaries of the binary, the histogram and points are both the missings and the complete values.

@@

The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, `vis_miss()` from \pkg{visdat} [@visdat]. This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:vis-miss)). The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B). Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation. Using \pkg{ggplot2} makes the plot easily customizable.

@@

- gg_miss_var and gg_miss_upset and vis_miss are all showing how we can summarise the missings.

@@

- but the addition of missing counts into a histogram, allows us to understand that there are many missings - and we have a skewed distribution o Ozone. But we want to know where these values might have been located. tying this back to the nabular structure, and also the terminology of MCAR MAR, etc. (We would expect if data were MCAR for these points to be representatively thoughout the histogram)
